---
title: "TCSI Adapt-Protect Score Analysis"
author: "Laurens Geffert"
date: "2022-12-01"
output: rmarkdown::github_document
---

<!-- forsys_tcsi_testing.md is generated from forsys_tcsi_testing.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '##',
  fig.retina = 2,
  fig.path = '../output/tcsi_a_p_score_analysis_files/tcsi_a_p_score_analysis')
```


```{r echo = FALSE, message = FALSE}
library(sf)
library(raster)
library(tidyverse)
library(magrittr)
```


Load data on Adapt and Protect scores and calculate an aggregation to speed up
the running of the code. Then calculate the Adapt-Protect-score from the union
of the two scores.

```{r data_load}
score_adapt <- raster('../data/TCSI/ecosystem/tif/adapt.tif')
score_protect <- raster('../data/TCSI/ecosystem/tif/protect.tif')

score_adapt <- aggregate(score_adapt, fact = 4, fun = mean)
score_protect <- aggregate(score_protect, fact = 4, fun = mean)

# Calculating opportunity score (a.k.a. Adapt-Protect score)
score_rescale <- function(raster1, raster2) {
  max_value <- max(raster1, raster2)
  rescale_factor <- (2 - sqrt(2)) / sqrt(2)
  raster_rescaled <- (2 * max_value + rescale_factor - 1) / (rescale_factor + 1)
  return(raster_rescaled)
}

score_opportunity <- score_rescale(score_adapt, score_protect)
```


Let's assume that we aim to treat 40% of the landscape. In this thought-
experiment we will try using in turn each of the Adapt score, the Protect score,
and the Adapt-Protect-score for finding the most important areas to treat.
The plots below show the cells we would select in each case.

```{r}
th_adapt <- quantile(score_adapt, .6)
th_protect <- quantile(score_protect, .6)
th_opportunity <- quantile(score_opportunity, .6)

stack(
  score_adapt > th_adapt,
  score_protect > th_protect,
  score_opportunity > th_opportunity) %>%
  set_names(c('adapt', 'protect', 'ap score')) %>%
  plot()
```


It looks like there is a lot of overlap. We will calculate what percentage of
cells in the Adapt-Protect-score case is selected because of their Adapt score
and what percentage is selected because of their Protect score.

```{r}
select_adapt <- ((score_opportunity > th_opportunity) & (score_adapt > th_adapt))
select_protect <- ((score_opportunity > th_opportunity) & (score_protect > th_protect))

n_cells_adapt <- select_adapt %>%
  values() %>%
  sum(na.rm = TRUE)
n_cells_protect <- select_protect %>%
  values() %>%
  sum(na.rm = TRUE)

calculate_proportion <- function(n1, n2, digits = 2) {
  {100 * n1 / (n1 + n2)} %>%
  round(digits = digits)
}

stack(
  select_adapt,
  select_protect) %>%
  set_names(c('selected for adapt', 'selected for protect')) %>%
  plot()
```

Comparing the areas selected using the Adapt-Protect-score to the areas
selected by **just** Adapt and Protect respectively, we can see that
`r calculate_proportion(n_cells_adapt, n_cells_protect)`
percent of cells are selected because of their Adapt score, while only 
`r calculate_proportion(n_cells_protect, n_cells_adapt)`
percent of cells are selected because of their Protect score!


