---
title: "ForSysR Tutorial"
author: "Laurens Geffert"
date: "2022-11-03"
output: rmarkdown::github_document
---

<!-- forsys_tutotial.md is generated from forsys_tutotial.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '##',
  fig.retina = 2,
  fig.path = '../output/forsys_tutotial_files/forsys_tutorial_')
```

## ForSysR Tutorial

Here we will provide a short example showing how the forsys package can be
used to build and solve simple multi-objective landscape management problems.
For brevity, we will use one of the built-in simulated datasets that is
distributed with the package. The code is based on the ForSysR tutorial from
the GitHub README.md but contains some modifications to the run parameters
and visualisations.

```{r setup, include=FALSE}
library(forsys)
library(sf)
library(tidyverse)
# for panel plots
library(cowplot)
```

Although forsys can support many different types of treatment unit data,
here our treatment units are represented as polygons in a spatial vector
format. Each polygon represents a different treatment unit.

```{r input_data, echo=FALSE}
# load treatment unit data
data(test_forest)
# show the first rows in the attribute table
glimpse(test_forest)
# project IDs look confusing when visualised, let's change their numbers
test_forest <- test_forest %>%
  inner_join(
    y = tibble(proj_id_old = 1:100, proj_id_new = sample.int(100, 100, replace = FALSE)),
    by = c('proj_id' = 'proj_id_old')) %>%
  select(-proj_id) %>%
  rename(proj_id = proj_id_new)
```

Let's visualize the data
```{r input_data_plot, echo=FALSE}
# plotting helper function
plot_test_forest <- function(sf, var_name) {
  ggp <- ggplot(sf, aes_string(fill = var_name)) +
    geom_sf(color = NA) +
    scale_fill_distiller(palette = 'Spectral') +
    theme_minimal() +
    theme(
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank())
  return(ggp)
}
# plot overview of the stands and and data
plot_vars <- c(
  'stand_id', 'proj_id', 'priority1', 'priority2', 'priority3',
  'priority4', 'threshold1', 'threshold2', 'ownership')
plot_grid(
  plot_test_forest(test_forest, plot_vars[1]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[2]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[3]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[4]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[5]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[6]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[7]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[8]) + theme(legend.position = 'none'),
  plot_test_forest(test_forest, plot_vars[9]) + theme(legend.position = 'none'),
  align = 'hv',
  axis = 'tblr',
  rel_heights = c(.8, .8, .8),
  labels = plot_vars[1:9],
  label_size = 6)
```


### 1 - Running a ForSys Scenario ----------------------------------------------

Forsys prioritizes projects by maximizing an objective given one or more
constraints. The objectives represent one or more management priorities while
the constraints may include a maximum cost or area treated. Thresholds are
environmental or categorical conditions that trigger the need to treat an
indiviudal treatment unit or stand (e.g., a particular ownership or minimum
forest cover). Forsys then builds projects and ranks them in order of their
priority. Projects can be either predefined units (e.g., watersheds) or can
be built dynamically.

Let’s set up a very simple forsys run to see how things work. We’ll use the
test_forest data shown above. We want to find the top 2000 ha within each
predefined project based on ‘priority1’.

```{r priority1, echo=FALSE}
plot_grid(
  plot_test_forest(test_forest, 'proj_id'),
  plot_test_forest(test_forest, 'priority1'))
```

We run forsys with the following arguments. Remember that these can also be
run using the json config file, as described above. Forsys always writes its
outputs to csv files saved within the output folder, but we can optionally
set it to write that data out to a list which has three elements containing
the outputs.

```{r run1, echo=FALSE}
stand_dat <- test_forest %>% st_drop_geometry()

run_outputs_1 <- forsys::run(
  return_outputs = TRUE,
  scenario_name = 'test_forest_run1_simple',
  stand_data = stand_dat,
  stand_id_field = 'stand_id',
  proj_id_field = 'proj_id',
  stand_area_field = 'area_ha',
  scenario_priorities = 'priority1',
  scenario_output_fields = c('area_ha', 'priority1', 'priority2', 'priority3', 'priority4'),
  proj_fixed_target =  TRUE,
  proj_target_field = 'area_ha',
  proj_target_value = 2000
)
```

Not surprisingly, the treatment rank of the projects selected corresponds
directly to those areas where are priority was highest, as plotted below.
Projeck rank #1 (darkest blue) is the highest ranked 

```{r run1_plot_projects, echo=FALSE}
test_forest %>%
  group_by(proj_id) %>%
  summarize() %>%
  left_join(
    y = select(run_outputs_1$project_output, proj_id, treatment_rank),
    by = 'proj_id') %>%
  ggplot(aes(fill = 100 - treatment_rank)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

Below we plot the stands rather than the project rank and only retain
those stands that were treated.

```{r run1_plot_stands, echo=FALSE}
test_forest %>%
  select(stand_id, proj_id) %>%
  inner_join(select(run_outputs_1$stand_output, stand_id)) %>%
  left_join(select(run_outputs_1$project_output, proj_id, treatment_rank)) %>%
  ggplot(aes(fill = 100 - treatment_rank)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

## 2 - Multiple priorities -----------------------------------------------------

Next we look at multiple priorities. Plotting priorities 1 and 2 shows that
areas where priority 1 are highest tend to be lower for priority 2.

```{r two_priorities, echo=FALSE}
test_forest %>%
  ggplot(aes(fill = priority1)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
test_forest %>%
  ggplot(aes(fill = priority2)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

Let’s see if forsys can find locations where we can achieve both objectives.
We prioritize on both variables, priority1 and priority2. We run forsys
weighting the two objectives from 0 to 5, which results in 21 scenarios.
We then filter the results to observe the outcome of the scenario where the
two objectives are equally weighted. The project rank graph represents areas
that are highest in both priorities.

```{r run2, echo=FALSE}
run_outputs_2 <- forsys::run(
  return_outputs = TRUE,
  scenario_name = 'test_forest_run2_scenarios',
  stand_data = stand_dat,
  stand_id_field = 'stand_id',
  proj_id_field = 'proj_id',
  stand_area_field = 'area_ha',
  scenario_priorities = c('priority1','priority2'),
  scenario_weighting_values = c('0 5 1'),
  scenario_output_fields = c('area_ha', 'priority1', 'priority2', 'priority3', 'priority4'),
  proj_fixed_target =  TRUE,
  proj_target_field = 'area_ha',
  proj_target_value = 2000
)

test_forest %>%
  group_by(proj_id) %>%
  summarize() %>%
  left_join(
    run_outputs_2$project_output %>%
    filter(Pr_1_priority1 == 1 & Pr_2_priority2 == 1) %>%
    select(proj_id, treatment_rank),
    by = 'proj_id') %>%
  ggplot(aes(fill = treatment_rank)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

## 3 - With Thresholds ---------------------------------------------------------

We expand on this scenario further by limiting stand selection by ownership
and threshold2. Any stands that don't meet the criteria are excluded.

```{r run3, echo=FALSE}
run_outputs_3 <- forsys::run(
  return_outputs = TRUE,
  scenario_name = 'test_forest_run3_threshold',
  stand_data = stand_dat,
  stand_id_field = 'stand_id',
  proj_id_field = 'proj_id',
  stand_area_field = 'area_ha',
  # this is the new parameter!
  stand_threshold = 'threshold2 == 1 & ownership == 2',
  scenario_priorities = 'priority1',
  scenario_weighting_values = c('0 5 1'),
  scenario_output_fields = c('area_ha', 'priority1', 'priority2', 'priority3', 'priority4'),
  proj_fixed_target =  TRUE,
  proj_target_field = 'area_ha',
  proj_target_value = 2000
)

test_forest %>%
  select(stand_id, proj_id) %>%
  inner_join(select(run_outputs_3$stand_output, stand_id), by = 'stand_id') %>%
  left_join(select(run_outputs_3$project_output, proj_id, treatment_rank), by = 'proj_id') %>%
  ggplot(aes(fill = 100 - treatment_rank)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

## 4 - Exploring different project prioritization methods ----------------------

Forsys can build projects dynamically using a package called Patchmax,
which requires some additional arguments.

```{r run4, echo=FALSE}
library(Patchmax)

# first we need to create an object describing stand adjacency and for gridded data, distance
adj <- Patchmax::calculate_adj(
  Shapefile = test_forest,
  St_id = test_forest$stand_id,
  # TODO: note, originally uses 'nb' but that doesn't work with large dataset
  method = 'buffer')
# TODO: This function takes quite a while to run!
# check inner workings of this function. Does it use actual spatial
# information and many-to-many relationship? Can we simplify this when all
# stands are rasters and use centroid distance?
dist <- Patchmax::calculate_dist(Shapefile = test_forest)

# then in the run functions we set the search distance weight to 10 to expand the search for high objective stands
run_outputs_4 <- forsys::run(
  return_outputs = TRUE,
  stand_data = stand_dat,
  scenario_name = 'test_forest_run4_patchmax',
  stand_id_field = 'stand_id',
  proj_id_field = 'proj_id',
  stand_area_field = 'area_ha',
  stand_threshold = 'threshold2 >= 1',
  scenario_priorities = 'priority2',
  scenario_output_fields = c('area_ha', 'priority1', 'priority2', 'priority3', 'priority4'),
  run_with_patchmax = TRUE,
  patchmax_stnd_adj = adj,
  patchmax_proj_size = 25000,
  patchmax_proj_number = 5,
  patchmax_st_distance = dist,
  # TODO: I assume this controls the tolerance for discontinuous treatment areas but not sure
  patchmax_SDW = 10
)

# plot the resulting project areas.
# TODO: project areas = treatment areas in this case?
test_forest %>%
  left_join(
    y = run_outputs_4$stand_output %>%
      # TODO: why is proj_id used as treatment rank here?
      mutate(treatment_rank = proj_id) %>%
      select(stand_id, treatment_rank),
    by = 'stand_id') %>%
  group_by(treatment_rank) %>%
  summarize() %>%
  ggplot(aes(fill = 5 - treatment_rank)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = 'Spectral') +
  theme_minimal()
```

